Realizing temporal graphs from fastest travel times

### 2023-03-21 New Paper/Realizing TG - extended.tex
starting writing some basic observations

\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate,anonymous]{lipics-v2021}
%This is a template for producing LIPIcs articles. 
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

@@

\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{mathrsfs}
\usepackage{wrapfig} %wrap the text around the picture

@@

\newcounter{guesscounter}

@@

@@

\begin{document}
\section{Ideas}
\begin{itemize}
    \item Symmetric distance matrix $D$: polytime?
    \item Use $|D|_\infty$ (maximum value in $D$) + max degree $\Delta$ as parameter $\rightarrow$ should give FPT
    \item FPT with feedback edge number
    \item use as parameter $\max_{i,j} |D_{i,j}-D_{j,i}|$
    \item approximation (additive or multiplicative), maybe with graph as input
\end{itemize}

@@

\todo[inline]{The matrix $D$ is m x m, and capturing the distances from an edge e to an edge f (zoom 14 Nov 22, GH).}
\todo[inline]{Check if the reduction works for the UPPER BOUNDED version (seems to work)}
\todo[inline]{how to extend the poly algorithms of cycles? Easy candidate: FES, FVS, series-parallel graphs, or tw=2 graphs}
\todo[inline]{FVS, $c$-factor approximation, hardness of approximation for $c<1.5$}
\end{document}

>>>>

\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
%This is a template for producing LIPIcs articles. 
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

@@

\usepackage{cleveref}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{mathrsfs}
\usepackage{wrapfig} %wrap the text around the picture

@@

@@

\newcommand{\deltaSymExactLong}{\textsc{Simple symmetric periodic Temporal Graph Realization}}
\newcommand{\deltaSymExact}{\textsc{Simple Symmetric TGR}}

@@

\begin{document}
	\maketitle

\problemdef{\deltaSymExactLong\ (\deltaSymExact)}
{An $n \times n$ symmetric matrix $D$, a positive integer $\Delta$.}
{Does there exist a graph $G=(V,E)$ with vertices $\{v_1,\ldots,v_{n}\}$ 
	and a $\Delta$-periodic labeling $\lambda: E \rightarrow \{1,2,\ldots,\Delta\}$ such that, 
	for every $i,j$, the duration of the fastest temporal path from $v_i$ to $v_j$ in the $\Delta$-periodic temporal graph $(G,\lambda,\Delta)$ is $D_{i,j}$?}

\problemdef{\kDeltaUpperBoundLong}
{A static graph $G=(V,E)$ with $n$ vertices, $n \times n$ matrix $D$ of positive integers and a bijective mapping $\phi$ from vertices $V$ of $G$ to the rows/columns of the matrix $D$.}
{Does there exist a $\Delta$-periodic labeling $\lambda: E \rightarrow \{1,2,\ldots,\Delta\}$ such that, 
	for every $i,j$, the duration of the fastest temporal path from $v_i$ to $v_j$ in the $\Delta$-periodic temporal graph $(G,\lambda,\Delta)$ is \emph{at most} $D_{\phi(i),\phi(j)}$.}

\section{Preliminaries and notation}
\todo[inline]{This section is copied from our paper}

An interval in $\mathbb N$ from $a$ to $b$ is denoted by $[a,b] = \{ i\in \mathbb N  :  a \leq i \leq b\}$; similarly, $[a] = [1,a]$.
An undirected graph~$G=(V,E)$ consists of a set~$V$ of vertices 
and a set~$E \subseteq V \times V$ of edges.
For a graph~$G$, we also denote by~$V(G)$ and~$E(G)$ the vertex and edge set of~$G$, respectively.
We denote an edge $e \in E$ between vertices $u,v \in V$ as a set $e=\{u,v\}$.
For the sake of simplicity of the representation, an edge $e$ is sometimes also denoted by $uv$. 
A path~$P$ in $G$ is a subgraph of $G$ with vertex set~$V(P)=\{v_1,\dots,v_k\}$ and edge set~$E(P)=\{\{v_i,v_{i+1}\} :  1\leq i<k\}$
(we often represent path~$P$ by the tuple~$(v_1,v_2,\dots,v_k)$).

Let $v_1,v_2,\ldots,v_n$ be the $n$ vertices of the graph $G$. 
For simplicity of the presentation~--and with a slight abuse of notation--~we refer during the paper to the entry $D_{i,j}$ of the matrix $D$ as 
$D_{a,b}$, where $a=v_i$ and $b=v_j$. That is, we put as indices of the matrix $D$ the corresponding vertices of $G$ whenever it is clear from the context.

We denote with $d(u,v,P)$ the duration of the fastest temporal path from $u$ to $v$ along path $P$. Note that $d(u,v,P) \geq d(u,v)$.

We remark that a fastest path between two vertices in a temporal graph can be computed in polynomial time~\cite{xuan_computing_2003,Wu2016Efficient}.
Hence, given a $\Delta$-periodic temporal graph $(G,\lambda,\Delta)$, we can compute in polynomial-time  the matrix $D$
which consists of durations of fastest temporal paths among all pairs of vertices in $(G,\lambda,\Delta)$.

\begin{definition}[Travel delays]
	Let $(G, \lambda)$ be a $\Delta$-periodic temporal graph.
	Let $e_1=uv$ and $e_2=vz$ be two incident edges in $G$ with $e_1 \cap e_2 = v$.
	We define the \emph{travel delay} from $u$ to $z$ at vertex $v$, denoted with $\tau_v^{uz}$,
	as the difference of the labels of $e_2$ and $e_1$, where we subtract the value of the label of $e_1$ from the label of $e_2$,  modulo~$\Delta$.
	More specifically:
	\begin{equation}\label{eq:def-VertexWaitingTime}
		\tau_v^{uz} \equiv \lambda (e_2) - \lambda(e_1) \pmod \Delta.
	\end{equation}
	Similarly, $\tau_v^{zu} \equiv \lambda (e_1) - \lambda(e_2) \pmod \Delta$.
\end{definition}
Intuitively, the value of $\tau_v^{uz}$ represents how long a temporal path waits at vertex $v$ when first taking edge $e_1=uv$ and then edge $e_2 = vz$.

\section{D is symmetric}
Before studying the problem in general let us first focus on two restricted cases.
\begin{theorem}
	If the input graph $G$ of \deltaSymExact\ is a clique or a $C_4$ then the problem is polynomial-time solvable. 
\end{theorem}

\begin{proof}
	We split the proof in two cases.

	First, when the input graph is a clique. In this case the desired labeling $\lambda$ can be arbitrary (i.e. $\lambda(e) \in [\Delta]$ for all $e \in E$), 
	as all vertices are on distance (and duration) $1$ from each other.
	Therefore 
	\begin{equation*}
		D_{i,j} =
		\begin{cases}
			0 & \text{if $i = j$}, \\
			1 & \text{else}
		\end{cases},
	\end{equation*}
	which is a symmetric matrix.

	Second, when the input graph $G$ is a $C_4$. Let $v_1, v_2, v_3, v_4$ be the vertices of $G$ with edges $e_1 = v_1 v_2, e_2 = v_2 v_3, e_3 = v_3 v_4, e_4 = v_4 v_1$.
	Now we set $\lambda(e_1) = \lambda (e_3) = \alpha$ and $\lambda(e_2) = \lambda(e_4) = \beta$, where w.l.o.g. $\alpha > \beta$.
	Let $\gamma = min \{ \alpha - \beta, \Delta + \beta - \alpha\}$,
	i.e. $\gamma$ is the minimum travel delay at vertex $v_i$ when traversing edges $e_{i-1}, e_i$, where indices are taken modulo $4$.
	We get that 
	\begin{equation*}
		D_{i,j} =
		\begin{cases}
			0 & \text{if $i = j$}, \\
			1 & \text{if $v_i$ and $v_j$ are neighbors},\\
			\gamma + 1 & \text{else}\\
		\end{cases}
	\end{equation*}
	Note, if $\alpha = \beta$ then the values in $D$ of non-neighboring vertices (i.e. vertices on distance $2$ from each other)
	are exactly $\Delta + 1$.
\end{proof}

Observe the following.
\begin{claim}
	Let $P = (u, v, w)$ be a path of length $2$, and let $(P,\lambda)$ be a $\Delta$-periodic temporal graph,
	 then $D_{u,w} = D_{w,v}$ iff $\tau_v^{uw} = \tau_v^{wu}$. 
\end{claim}
\begin{proof}
	Let us denote with $\alpha = \lambda(uv)$ and  $\beta = \lambda(vw)$,
	w.l.o.g. we suppose also that $\alpha > \beta$.
	Then $d(u,w) = \Delta + \beta - \alpha + 1$ 
	and $d(w,u) = \alpha - \beta + 1$.
	By the definition of travel delays we get $d(u,w) = \tau_v^{uw} + 1$ and $d(w,u) = \tau_v^{wu} + 1$. 
	Since $D_{u,w} = D_{w,v}$ we get that $\tau_v^{uw} = \tau_v^{wu}$.
\end{proof}

From the above claim we get the following.

\begin{claim}
	Let $P = (u, v, w)$ be a path of length $2$, and let $(P,\lambda)$ be a $\Delta$-periodic temporal graph, 
	then $D_{u,w} = D_{w,v}$
	iff
	$\lambda(e) \in \{\frac{\Delta}{2}, \Delta\}$, if $\Delta$ is even and
	$\lambda(e) \in \{\Delta\}$, if $\Delta$ is odd,
	for all $e \in E(P)$.
\end{claim}

\begin{proof}
		Let us denote with $\alpha = \lambda(uv)$ and  $\beta = \lambda(vw)$,
		w.l.o.g. we suppose also that $\alpha > \beta$.
		Since
		$\tau_v^{uw} = \tau_v^{wu}$
		we get that
		\begin{equation*}
			\Delta + \beta - \alpha \equiv \alpha - \beta \pmod\Delta,
		\end{equation*}
		which is equivalent to
		\begin{equation}\label{eq:sol-delta}
			\Delta  \equiv 2 (\alpha - \beta) \pmod\Delta.
		\end{equation}
		Since $\alpha > \beta$, the only possible solution of \cref{eq:sol-delta} is when $\beta \equiv \frac{\alpha}{2} \pmod \Delta$, which exists only in the case when $\Delta$ is even.
		If we now allow $\alpha = \beta$ then we get another solution of \cref{eq:sol-delta}.

		W.l.o.g. we can now suppose that $\alpha = \Delta$.
		If $\Delta$ is odd, we have a unique solution with the value $\alpha=\beta = \Delta$,
		while in the case when $\Delta$ is odd we get that $\alpha, \beta \in \{\Delta, \frac{\Delta}{2}\}$. 
\end{proof}

We can extend the above result to all graphs as we want to have a periodic labeling of a graph $G$, that gives rise to a symmetric matrix $D$. 
Then the above has to hold for all pairs of vertices on distance $2$.
Therefore we get the following two results.

\begin{claim}\label{claim:symmetric-odd}
	When $\Delta$ is odd, the only possibility of the matrix $D$, of all durations, being symmetric is when all labels of $G$ are the same.
\end{claim}

\begin{claim}\label{claim:symmetric-even}
	When $\Delta$ is even, the only possibility of the matrix $D$, of all durations, being symmetric is when all labels of $G$ are from $\{\frac{\Delta}{2}, \Delta\}$.
\end{claim}

From all of the above we get the following result.
\begin{theorem}
	\deltaSymExact\ can be solved in polynomial time.
\end{theorem}

\begin{proof}
	TODO
\end{proof}

\section{Upper bounds in the matrix D}

\clearpage

\section{Ideas}
\begin{itemize}
    \item Symmetric distance matrix $D$: polytime?
    \item Use $|D|_\infty$ (maximum value in $D$) + max degree $\Delta$ as parameter $\rightarrow$ should give FPT
    \item FPT with feedback edge number
    \item use as parameter $\max_{i,j} |D_{i,j}-D_{j,i}|$
    \item approximation (additive or multiplicative), maybe with graph as input
\end{itemize}

@@

\todo[inline]{The matrix $D$ is m x m, and capturing the distances from an edge e to an edge f (zoom 14 Nov 22, GH).}
\todo[inline]{Check if the reduction works for the UPPER BOUNDED version (seems to work)}
\todo[inline]{how to extend the poly algorithms of cycles? Easy candidate: FES, FVS, series-parallel graphs, or tw=2 graphs}
\todo[inline]{FVS, $c$-factor approximation, hardness of approximation for $c<1.5$}

\bibliography{../bibliography}	
\end{document}

### 2023-03-21 arxiv.tex
starting writing some basic observations

    From the above definition and the definition of the duration of the temporal path $P$ we get the following two observations.
    \begin{observation}\label{obs:durationPwithWaitingTimes}
        Let $P = (v_0, v_1, \dots, v_p)$ be the underlying path of the temporal path
        $(P, \lambda)$ from $v_0$ to $v_p$.
        Then $d(P,\lambda) = \sum_{i = 1}^{p-1} \tau_{v_i}^{v_{i-1}v_i} + 1 $.
    \end{observation}
    \begin{proof}
        For the simplicity of the proof denote $t_i = \lambda(v_{i-1}v_i)$, and suppose that $t_i \leq t_{i+1}$, for all $i \in \{1,2,3,\dots,p\}$.
        Then
        \begin{align*}
        \sum_{i = 1}^{p-1} \tau_{v_i}^{v_{i-1}v_i} + 1  
        &= \sum _{i = 1}^{p-1} (t_{i+1} - t_i) + 1 \\
        & = (t_2 - t_1) + (t_3  t_2) + \cdots + (t_p - t_{p-1}) + 1  \\
        & = t_{p-1} - t_1 + 1\\
        & = d(P)
        \end{align*}
        Now in the case when $t_i > t_{i+1}$ we get that $\tau_{v_i}^{v_{i-1}v_{i+1}} = \Delta + t_{i+1} - t_i$.
        At the end this still results in the correct duration as the last time we traverse the path $P$ is not exactly $t_p$ but $k \lambda + t_p$, for some $k$.
    \end{proof}
    We also get the following.
    \begin{observation}\label{obs:travel-delays-both-directions}
    Let $(G, \lambda)$ be a temporal graph satisfying conditions of problem \deltaExact.
    For any two incident edges $e_1 = uv$ and $e_2 = vz$ on vertices $u,v,z \in V$, with $e_1 \cap e_2 = v$, we have $\tau_v^{zu} = \Delta - \tau_v^{uz} \pmod \Delta$.
    \end{observation}

@@

\subsection{Polynomial-time algorithm for trees}\label{sec:treealgo}
%All of the above is used in the design of the following polynomial-time algorithms.
We are now ready to provide a polynomial-time algorithm for \deltaExact\ when the underlying graph is a tree.
Let $D$ be a matrix from \deltaExact\ and let the underlying graph $G$ of $D$ be a tree on $n$ vertices $\{v_1, v_2, \dots, v_n\}$.
Let $v_i,v_j$ be two arbitrary vertices in $G$. Then we know that there exists a unique (static) path $P$ among them.
Consequently, it follows that the temporal paths $(P_{i,j},\lambda)$ from $v_i$ to $v_j$ and $(P_{j,i},\lambda)$ from $v_j$ to $v_i$ are also unique, up to modulo of the period $\Delta$ of the labeling $\lambda$,
and therefore are the fastest.
Then $D$ is of the following form:
\begin{equation*}
    D_{i,j} =
    \begin{cases}
    0 & \text{if $i = j$}, \\
    1 & \text{if $v_i$ and $v_j$ are neighbours in G}\\
    d(P_{i,j},\lambda) & \text{else}
    \end{cases},
\end{equation*}
where $d(P_{i,j},\lambda)$ is the duration of the (unique) temporal path $(P_{i,j},\lambda)$ from $v_i$ to $v_j$.

@@

\begin{figure}[t]
	\centering
	\begin{tikzpicture}[scale=.85]
		%%%S_uv
		\node[vert,label=below:$u{=} v_1$] (v1) at (1,0) {};
		\node[vert, label=below:$v_2$] (v2) at (2,0) {};
		\node[vert,label=below:$v_3$] (v3) at (3,0) {};
		\node[vert,label=below:$v_{i-1}$] (vi-1) at (6,0) {};
		\node[vert,label=below:$v_i$] (vi) at (7,0) {};
		\node[vert,label=below:$v_{i+1}$] (vi+1) at (8,0) {};
		\node[vert,label=below:$v_{p} {=} v$] (v) at (10,0) {};

		\draw (v1) -- node[label={above:$t_{v_2}$},yshift=-5] {} (v2) -- (v3) -- (vi-1) -- node[label={above:$e$},yshift=-5] {} (vi) -- node[label={above:$f$},yshift=-5] {} (vi+1) -- (v);

		\draw (v1) to [out=80,in=100,distance=3cm]
		node[pos=0.1,label={left:$t_{u_2}$}] {}
		node[vert,pos=0.15,label=above:$u_2$] {}
		node[vert,pos=0.3,label=above:$u_3$] {}
		node[vert,pos=0.85,label=above:$u_{j-1}$] {}
		node[pos=0.9,label={right:$g$}] {} (vi);

		%\draw[transform canvas={yshift=0mm}, blue]
		%(v1) -- (v2) -- (v3) -- node[pos=0.3,yshift=-2,label=above:$P_{u,v}$] {} (vi-1) -- (vi) ;

		%\draw[red]
		%(v1) edge[diredge2] [out=85,in=95,distance=2.1cm] node[pos=0.3,yshift=2,label=above:$P_{u,w}$] {} (v3) ;
	\end{tikzpicture}
	\caption{An example of the situation in \cref{lemma:FPT-uv-LabelAlmostalledges},
		where we assume that the fastest temporal path from $u$ to $v$ is $P_{u,v} = (u=v_1, v_2, \dots v_p)$,
		and the fastest temporal path from $u$ to $v_i$ in $P_{u,v}$ is $P = (u, u_2, u_3, \dots, v_i)$.
		We denote with $Q = (u=v_1, v_2, \dots, v_i)$ and with $R = (v_i, v_{i+1}, \dots, v_p=v)$.
		\label{fig:FPT-uv-Labelalledges}}
\end{figure}

>>>>

    From the above definition and the definition of the duration of the temporal path $P$ we get the following two observations.
    \begin{observation}\label{obs:durationPwithWaitingTimes}
        Let $P = (v_0, v_1, \dots, v_p)$ be the underlying path of the temporal path
        $(P, \lambda)$ from $v_0$ to $v_p$.
        Then $d(P,\lambda) = \sum_{i = 1}^{p-1} \tau_{v_i}^{v_{i-1}v_i} + 1 $.
    \end{observation}
    \begin{proof}
        For the simplicity of the proof denote $t_i = \lambda(v_{i-1}v_i)$, and suppose that $t_i \leq t_{i+1}$, for all $i \in \{1,2,3,\dots,p\}$.
        Then
        \begin{align*}
        \sum_{i = 1}^{p-1} \tau_{v_i}^{v_{i-1}v_i} + 1  
        &= \sum _{i = 1}^{p-1} (t_{i+1} - t_i) + 1 \\
        & = (t_2 - t_1) + (t_3 - t_2) + \cdots + (t_p - t_{p-1}) + 1  \\
        & = t_{p-1} - t_1 + 1\\
        & = d(P)
        \end{align*}
        Now in the case when $t_i > t_{i+1}$ we get that $\tau_{v_i}^{v_{i-1}v_{i+1}} = \Delta + t_{i+1} - t_i$.
        At the end this still results in the correct duration as the last time we traverse the path $P$ is not exactly $t_p$ but $k \lambda + t_p$, for some $k$.
    \end{proof}
    We also get the following.
    \begin{observation}\label{obs:travel-delays-both-directions}
    Let $(G, \lambda)$ be a temporal graph satisfying conditions of problem \deltaExact.
    For any two incident edges $e_1 = uv$ and $e_2 = vz$ on vertices $u,v,z \in V$, with $e_1 \cap e_2 = v$, we have $\tau_v^{zu} = \Delta - \tau_v^{uz} \pmod \Delta$.
    \end{observation}

@@

\subsection{Polynomial-time algorithm for trees}\label{sec:treealgo}
%All of the above is used in the design of the following polynomial-time algorithms.
We are now ready to provide a polynomial-time algorithm for \deltaExact\ when the underlying graph is a tree.
Let $D$ be a matrix from \deltaExact\ and let the underlying graph $G$ of $D$ be a tree on $n$ vertices $\{v_1, v_2, \dots, v_n\}$.
Let $v_i,v_j$ be two arbitrary vertices in $G$. Then we know that there exists a unique (static) path $P$ among them.
Consequently, it follows that the temporal paths $(P_{i,j},\lambda)$ from $v_i$ to $v_j$ and $(P_{j,i},\lambda)$ from $v_j$ to $v_i$ are also unique, up to modulo of the period $\Delta$ of the labeling $\lambda$,
and therefore are the fastest.
Then $D$ is of the following fo\rm:
\begin{equation*}
    D_{i,j} =
    \begin{cases}
    0 & \text{if $i = j$}, \\
    1 & \text{if $v_i$ and $v_j$ are neighbours in G}\\
    d(P_{i,j},\lambda) & \text{else}
    \end{cases},
\end{equation*}
where $d(P_{i,j},\lambda)$ is the duration of the (unique) temporal path $(P_{i,j},\lambda)$ from $v_i$ to $v_j$.

@@

\begin{figure}[t]
	\centering
	\includegraphics{fig-lemma}
	\caption{An example of the situation in \cref{lemma:FPT-uv-LabelAlmostalledges},
		where we assume that the fastest temporal path from $u$ to $v$ is $P_{u,v} = (u=v_1, v_2, \dots v_p)$,
		and the fastest temporal path from $u$ to $v_i$ in $P_{u,v}$ is $P = (u, u_2, u_3, \dots, v_i)$.
		We denote with $Q = (u=v_1, v_2, \dots, v_i)$ and with $R = (v_i, v_{i+1}, \dots, v_p=v)$.
		\label{fig:FPT-uv-Labelalledges}}
\end{figure}
